---
title: "NeuroFlow Data Challenge"
author: "Ty Painter"
date: "`r date()`"
output: github_document
---

```{r}
# load packages
library(tidyverse)
library(ggplot2)
library(dplyr)
```

# Load Data
```{r}
library(readr)
phq_all_final <- read_csv("/Users/TyPainter1/Downloads/neuroflow_data_challenge/phq_all_final.csv") # update file path to run on local machine
head(phq_all_final)
```

```{r}
summary(phq_all_final)
```

# Exploratory Data Analysis
## Categorize Scores into Severity Label

```{r}
phq_all_final$severity <- cut(phq_all_final$score,c(-1,5,10,15,21))
levels(phq_all_final$severity) = c("Low to Minimal", "Mild", "Moderate", "Severe")
head(phq_all_final)
```
### Display Raw Counts of Severity Labels
```{r}
ggplot(phq_all_final, aes(severity)) +
  geom_histogram(stat="count") +
    labs(
    title = "Distribution of Severity Labels",
    x = "Severity",
    y = "# of Assessments",
    fill = "Severity"
  ) +
  scale_y_continuous(labels = function(y){paste0(y/1000, 'K')}) 
```

### Display Proportions of Severity Labels
```{r}
ggplot(phq_all_final, aes(x = severity)) +  
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    labs(
    title = "Proportion of Severity Labels",
    x = "Severity",
    y = "% of Assessments",
    fill = "Severity"
  ) +
  scale_y_continuous(labels = function(y){paste0(y*100, '%')}) 
```

## Evaluate Progress Over Time
```{r}
# create a visit number column for each patient
visit_num = rep(NA,nrow(phq_all_final))
visit_num[1] = 1
for (j in 2:nrow(phq_all_final)) {
  i = j-1
  if (phq_all_final$patient_id[i] == phq_all_final$patient_id[j]) {
    visit_num[j] = visit_num[i] + 1
  } else {
    visit_num[j] = 1
  }
}

phq_all_final <- phq_all_final %>% 
  mutate(visit_num = visit_num)
```

### Raw Count of Visits per Patient
```{r}
phq_all_final %>% 
  group_by(patient_id) %>% 
  summarize(total_visits = max(visit_num)) %>% 
  ggplot(aes(total_visits)) +
  geom_histogram(stat="count") +
    labs(
    title = "Distribution of Number of Visits per Patient",
    x = "# of Visits",
    y = "# of Patients"
  ) +
  scale_y_continuous(labels = function(y){paste0(y/1000, 'K')}) 
```

### Proportions of Visits per Patient 
```{r}
phq_all_final %>% 
  group_by(patient_id) %>% 
  summarize(total_visits = max(visit_num)) %>% 
  ggplot(aes(x = total_visits)) +  
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  labs(
    title = "Proportion of Visits per Patients",
    x = "# of Visits",
    y = "% of Patients"
  ) +
  scale_y_continuous(labels = function(y){paste0(y*100, '%')}) 
```

### Average GAD7 Scores per Number of Patient Visits
```{r}
phq_all_final %>% 
  group_by(visit_num) %>% 
  summarize(avg_score = mean(score)) %>% 
  ggplot(aes(x = visit_num, y = avg_score)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed", col="red") +
  geom_text(aes(6,0,label="Low to Minimal", vjust=-3)) +
  geom_hline(yintercept=6, linetype="dashed", col="red") +
  geom_text(aes(5,6,label="Mild", vjust=-3)) +
  geom_hline(yintercept=11, linetype="dashed", col="red") +
  geom_text(aes(5,11,label="Moderate", vjust=-3)) +
  geom_hline(yintercept=16, linetype="dashed", col="red") +
  geom_text(aes(5,16,label="Severe", vjust=-3)) +
  labs(
    title = "Average GAD7 Scores per Number of Patient Visits",
    x = "# of Visits",
    y = "GAD7 Score")
```

### Median GAD7 Score per Number of Patient Visits
```{r}
phq_all_final %>% 
  group_by(visit_num) %>% 
  summarize(avg_score = median(score)) %>% 
  ggplot(aes(x = visit_num, y = avg_score)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed", col="red") +
  geom_text(aes(6,0,label="Low to Minimal", vjust=-1)) +
  geom_hline(yintercept=6, linetype="dashed", col="red") +
  geom_text(aes(5,6,label="Mild", vjust=-3)) +
  geom_hline(yintercept=11, linetype="dashed", col="red") +
  geom_text(aes(5,11,label="Moderate", vjust=-3)) +
  geom_hline(yintercept=16, linetype="dashed", col="red") +
  geom_text(aes(5,16,label="Severe", vjust=-3)) +
  labs(
    title = "Median GAD7 Scores per Number of Patient Visits",
    x = "# of Visits",
    y = "GAD7 Score")
```

# Final Analysis
**What insights can you draw?**

The main insight I drew from my analysis is the patient visits/assessments seem to be working because 90% of patients are making only 7 visits or less and both the median and average GAD7 scores are in the Low to Minimal risk range. However, there seems to be a sharp increase in GAD7 scores around 40 visits. This is logical since a higher GAD7 score would require more visits which skews the graph, but it seems that once a patients goes over 40 visits diminishing returns start to begin.
```{r}
total_visits <- phq_all_final %>% 
  group_by(patient_id) %>% 
  summarize(total_visits = max(visit_num)) 
quantile(total_visits$total_visits, .9)
```

**What assumptions have you made about the data?**

I did not make many assumptions about the data because everything was pretty self explanatory. The main assumption I made was that the data was entered correctly which I checked in the begging with the summary command and nothing irregular stood out.

**What are 2-3 additional pieces of information that would be important to collect?**

Additional pieces of information I wish I had was the type of diagnosis or treatment that each patient was associated with. I think this would allow for a better evaluation based on the diagnosis/treatment. Another piece of information I would like to look at would be the patient responses to each specific question of the GAD7. Again, I think this could provide further insight into particular questions that relate to patient struggles. Also, if I had more time to complete my analysis I would have liked to look into the length of time between patient assessments and how that affects their GAD7 scores.