---
title: "NeuroFlow Data Challenge"
author: "Ty Painter"
date: "`r date()`"
output: github_document
---

```{r}
# load packages
library(tidyverse)
library(ggplot2)
library(dplyr)
```

# Load Data
```{r}
library(readr)
phq_all_final <- read_csv("/Users/TyPainter1/Downloads/neuroflow_data_challenge/phq_all_final.csv") # update file path to run on local machine
head(phq_all_final)
```

```{r}
summary(phq_all_final) # check variable values
```

# Exploratory Data Analysis
## Categorize Scores into Severity Label

```{r}
phq_all_final$severity <- cut(phq_all_final$score,c(-1,5,10,15,21)) # set ranges for severity labels
levels(phq_all_final$severity) = c("Low to Minimal", "Mild", "Moderate", "Severe")
head(phq_all_final)
```
### Raw Counts of Severity Labels
```{r}
ggplot(phq_all_final, aes(severity)) +
  geom_histogram(stat="count") +
    labs(
    title = "Distribution of Severity Labels",
    x = "Severity",
    y = "# of Assessments",
    fill = "Severity"
  ) +
  scale_y_continuous(labels = function(y){paste0(y/1000, 'K')}) 
```

### Proportions of Severity Labels
```{r}
ggplot(phq_all_final, aes(x = severity)) +  
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    labs(
    title = "Proportion of Severity Labels",
    x = "Severity",
    y = "% of Assessments",
    fill = "Severity"
  ) +
  scale_y_continuous(labels = function(y){paste0(y*100, '%')}) 
```

## Evaluate Progress Over Time
```{r}
# create a assessment number column for each patient
assessment_num = rep(NA,nrow(phq_all_final))
assessment_num[1] = 1 # initiate first patient 
for (j in 2:nrow(phq_all_final)) {
  i = j-1
  if (phq_all_final$patient_id[i] == phq_all_final$patient_id[j]) {
    assessment_num[j] = assessment_num[i] + 1 # if same patient, add to assessment total
  } else {
    assessment_num[j] = 1 # if different patient, start new count
  }
}

phq_all_final <- phq_all_final %>% 
  mutate(assessment_num = assessment_num) # add assessment number to original df
```

### Distribution of Assessments per Patient
```{r}
phq_all_final %>% 
  group_by(patient_id) %>% 
  summarize(total_assessments = max(assessment_num)) %>% # take the largest assessment num of each patient for total assessments for each patient
  ggplot(aes(total_assessments)) +
  geom_histogram(stat="count") +
    labs(
    title = "Distribution of Number of Assessments per Patient",
    x = "# of Assessments",
    y = "# of Patients"
  ) +
  scale_y_continuous(labels = function(y){paste0(y/1000, 'K')}) 
```

### Proportions of Assessments per Patient 
```{r}
phq_all_final %>% 
  group_by(patient_id) %>% 
  summarize(total_assessments = max(assessment_num)) %>% # take the largest assessment num of each patient for total assessments for each patient
  ggplot(aes(x = total_assessments)) +  
  geom_bar(aes(y = (..count..)/sum(..count..))) + # calculate proportion
  labs(
    title = "Proportion of Assessments per Patients",
    x = "# of Assessments",
    y = "% of Patients"
  ) +
  scale_y_continuous(labels = function(y){paste0(y*100, '%')}) 
```

### Average GAD7 Scores per Number of Patient Assessments
```{r}
phq_all_final %>% 
  group_by(assessment_num) %>% 
  summarize(avg_score = mean(score)) %>% # avg score per # of assessments
  ggplot(aes(x = assessment_num, y = avg_score)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed", col="red") + # add severity labels
  geom_text(aes(6,0,label="Low to Minimal", vjust=-3)) +
  geom_hline(yintercept=6, linetype="dashed", col="red") +
  geom_text(aes(5,6,label="Mild", vjust=-3)) +
  geom_hline(yintercept=11, linetype="dashed", col="red") +
  geom_text(aes(5,11,label="Moderate", vjust=-3)) +
  geom_hline(yintercept=16, linetype="dashed", col="red") +
  geom_text(aes(5,16,label="Severe", vjust=-3)) +
  labs(
    title = "Average GAD7 Scores per Number of Patient Assessments",
    x = "# of Assessments",
    y = "GAD7 Score")
```

### Median GAD7 Score per Number of Patient Assessments
```{r}
phq_all_final %>% 
  group_by(assessment_num) %>% 
  summarize(avg_score = median(score)) %>% 
  ggplot(aes(x = assessment_num, y = avg_score)) + # avg score per # of assessments
  geom_line() +
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed", col="red") + # add severity labels
  geom_text(aes(6,0,label="Low to Minimal", vjust=-1)) +
  geom_hline(yintercept=6, linetype="dashed", col="red") +
  geom_text(aes(5,6,label="Mild", vjust=-3)) +
  geom_hline(yintercept=11, linetype="dashed", col="red") +
  geom_text(aes(5,11,label="Moderate", vjust=-3)) +
  geom_hline(yintercept=16, linetype="dashed", col="red") +
  geom_text(aes(5,16,label="Severe", vjust=-3)) +
  labs(
    title = "Median GAD7 Scores per Number of Patient Assessments",
    x = "# of Assessments",
    y = "GAD7 Score")
```

# Final Analysis
**What insights can you draw?**

  It seems that 3-6 assessments tend to result in the lowest CAD7 scores which is good since 88% of patients participate in 6 assessments or less. However, there is cause for concern where 34% of patients took an initial assessment and scored 10 or greater but did not take another assessment although a score of 10+ indicates further evaluation. 

```{r}
total_assessments <- phq_all_final %>% 
  group_by(patient_id) %>% 
  summarize(total_assessments = max(assessment_num)) 
quantile(total_assessments$total_assessments, .88) # find 88th quantile
```

```{r}
phq_all_final <- left_join(phq_all_final, total_assessments, by = "patient_id") # merge total assessments for each patient to original df

no_return <- phq_all_final %>% 
  filter(total_assessments == 1 & score >= 10) %>% # patients that scored 10+ on initial assessment and did not return
  nrow()

return <- phq_all_final %>% 
  filter(assessment_num == 1 & score >= 10) %>% # patients that scored 10+ on initial assessment 
  nrow()

no_return / return
```

  There seems to be a sharp increase in GAD7 scores once a patient surpasses assessments in the mid 30s. This is logical since a higher GAD7 score would require more assessments, but it seems that once a patients passes 35 assessments diminishing returns start to begin. One thing to note about this analysis is that on 8 out of 15,502 patients had over 30 assessments which heavily skews the graph.

```{r}
total_assessments %>% 
  filter(total_assessments > 30) %>% # number of patients with over 30 assessments
  nrow()
```

**What assumptions have you made about the data?**

I did not make many assumptions about the data because everything was pretty self explanatory. The main assumption I made was that the data was entered correctly which I checked in the begging with the summary command and nothing irregular stood out.

**What are 2-3 additional pieces of information that would be important to collect?**

Additional pieces of information I wish I had was the type of diagnosis or treatment that each patient was associated with. I think this would allow for a better analysis based on the diagnosis/treatment. Another piece of information I would like to look at would be the patient responses to each specific question of the GAD7. Again, I think this could provide further insight into particular questions that relate to patient struggles. Also, if I had more time to complete my analysis I would have liked to look into the length of time between patient assessments and how that affects their GAD7 scores.



